A Sound Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2025-08-21 paper/COLA/paper-sblp.tex
changing authors.
Signed-off-by: Rodrigo Ribeiro <rodrigogribeiro@gmail.com>

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

@@

\author{Miguel Nunes}
\affiliation{\institution{Santa Catarina State University}\country{}}
\email{miguel.alfredo.nunes@gmail.com}

@@

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{da Silveira et al.}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Modal logics, soundness, Coq proof assistant}

@@

However, writing a proof term whose type is that of a logical formula
can be challenging, even for very simple propositions.  In order to
make the writing of complex proofs easier, Coq provides
\emph{tactics}, which are commands that can be used to construct proof
terms in a more user-friendly way.

As a tiny example, consider the task of proving the following 
formula of the propositional logic:
\[
(A \to B)\to (B\to C) \to A \to C
\]
In Coq, such theorem can be expressed as:
\begin{minted}{coq}
Section EXAMPLE.
   Variables A B C: Prop.
   Theorem example: (A -> B) -> (B -> C) -> A -> C.
   Proof.
       intros H H' HA.
       apply H'.
       apply H.
       assumption. 
   Qed.
End EXAMPLE.
\end{minted}
In the previous source code piece, we have defined a Coq section named
\mintinline{coq}{EXAMPLE}\footnote{In Coq, we can use sections to delimit the
  scope of local variables.} which declares variables \mintinline{coq}{A},
\mintinline{coq}{B} and \mintinline{coq}{C} as being propositions (i.e. with type
\mintinline{coq}{Prop}). Tactic \mintinline{coq}{intros} introduces variables
\mintinline{coq}{H}, \mintinline{coq}{H'} and \mintinline{coq}{HA} into the (typing) context,
respectively with types \mintinline{coq}{A -> B}, \mintinline{coq}{B -> C} and
\mintinline{coq}{A} and leaves goal \mintinline{coq}{C} to be proved. Tactic
\mintinline{coq}{apply}, used with a term \mintinline{coq}{t}, generates goal
\mintinline{coq}{P}
when there exists \mintinline{coq}{t: P -> Q} in the typing context and the
current goal is \mintinline{coq}{Q}. Thus, \mintinline{coq}{apply H'} changes the goal
from \mintinline{coq}{C} to \mintinline{coq}{B} and \mintinline{coq}{apply H}
changes the goal to \mintinline{coq}{A}. Tactic \mintinline{coq}{assumption}
traverses the typing context to find a hypothesis that matches the goal.

We define next a proof of the previous propositional logical formula
that, in contrast to the previous proof, that was built using tactics
(\mintinline{coq}{intros}, \mintinline{coq}{apply} and \mintinline{coq}{assumption}), is coded
directly as a function:
\begin{minted}{coq}
Definition example: (A -> B) -> (B -> C) -> A -> C :=
  fun (H: A -> B) (H': B -> C) (HA: A) => H' (H HA).
\end{minted}
However, even for very simple theorems, coding a definition directly
as a Coq term can be a hard task. Because of this, the use of tactics
has become the standard way of proving theorems in Coq. Furthermore,
the Coq proof assistant provides not only a great number of tactics
but also a domain-specific language for scripted proof automation,
called $\mathcal{L}$tac. Details about $\mathcal{L}$tac and Coq can be found
in~\cite{Chlipala13,Bertot10,manual_coq}.

>>>>

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Miguel Nunes}
\affiliation{\institution{Santa Catarina State University}\country{}}
\email{miguel.alfredo.nunes@gmail.com}

@@

@@

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Nunes et al.}

@@

\begin{CCSXML}
  <ccs2012>
    <concept>
        <concept_id>10003752.10003790.10003793</concept_id>
        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
        <concept_significance>500</concept_significance>
        </concept>
    <concept>
        <concept_id>10003752.10003790.10002990</concept_id>
        <concept_desc>Theory of computation~Logic and verification</concept_desc>
        <concept_significance>500</concept_significance>
        </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Modal and temporal logics}
\ccsdesc[500]{Theory of computation~Logic and verification}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Modal logic, soundness, Coq proof assistant}

@@

In order to make writing proofs easier, Coq provides
\emph{tactics}, which are commands that can be used to construct proof
terms in a more user-friendly way. As an example, consider the task of proving the following
formula of the propositional logic:
\[
(A \to B)\to (B\to C) \to A \to C
\]
In Coq, such theorem can be expressed as:
\begin{minted}{coq}
Section EXAMPLE.
  Variables A B C: Prop.
  Theorem example: (A -> B) -> (B -> C) -> A -> C.
  Proof.
    intros H H' HA.
    apply H'.
    apply H.
    assumption.
  Qed.
End EXAMPLE.
\end{minted}
In the above, we have defined a Coq section named
\mintinline{coq}{EXAMPLE}\footnote{In Coq, we can use sections to delimit the
  scope of local variables.} which declares variables \mintinline{coq}{A},
\mintinline{coq}{B} and \mintinline{coq}{C} as being propositions (i.e. with type
\mintinline{coq}{Prop}). Tactic \mintinline{coq}{intros} introduces variables
\mintinline{coq}{H}, \mintinline{coq}{H'} and \mintinline{coq}{HA} into the typing context,
respectively with types \mintinline{coq}{A -> B}, \mintinline{coq}{B -> C}, and
\mintinline{coq}{A}, and leaves goal \mintinline{coq}{C} to be proved. Tactic
\mintinline{coq}{apply}, used with a term \mintinline{coq}{t}, generates goal
\mintinline{coq}{P}
when there exists \mintinline{coq}{t: P -> Q} in the typing context and the
current goal is \mintinline{coq}{Q}. Thus, \mintinline{coq}{apply H'} changes the goal
from \mintinline{coq}{C} to \mintinline{coq}{B} and \mintinline{coq}{apply H}
changes the goal to \mintinline{coq}{A}. Tactic \mintinline{coq}{assumption}
traverses the typing context to find a hypothesis that matches the goal,
concluding the proof.

We define next a proof of the previous propositional logical formula
that, in contrast to the previous proof, that was built using tactics, is coded
directly as a function:
\begin{minted}{coq}
Definition example: (A -> B) -> (B -> C) -> A -> C :=
  fun (H: A -> B) (H': B -> C) (HA: A) => H' (H HA).
\end{minted}
However, even for very simple theorems, coding a definition directly
as a Coq term can be a hard task. Because of this, the use of tactics
has become the standard way of proving theorems in Coq. Furthermore,
the Coq proof assistant provides not only a great number of tactics
but also a domain-specific language for scripted proof automation,
called $\mathcal{L}$tac. Details about $\mathcal{L}$tac and Coq can be found
in~\cite{Chlipala13,Bertot10,manual_coq}.
